Composable Concurrency Models

### 2016-11-30 paper/paper.tex
Did another round of revisions; unsuccessfully attempted to compile a list of things worse than BibTeX

Given that an application is likely to make use of more than one concurrency model, programmers would prefer that different types of models could safely interact. However, different models do not necessarily work well together, nor are they designed to. Recent work has attempted to identify common ``building blocks'' that could be used to compose a variety of models, possibly eliminating subtle problems when different models interact and allowing models to be represented at lower levels without resorting to rough approximations~\cite{Marr2009, Marr2012, Swalens2014, Ziv2015}.

@@

Figure~\ref{figure:single} illustrates a simple program in which a single thread reads ($=$) and writes ($\leftarrow$) values on a variable $x$, denoted by a segmented bar. The program satisfies an intuitive model of how variables should behave---each time $x$ is read, the value returned is equal to the value most recently written. Because the thread has exclusive control of $x$, that assumption is valid.

\begin{figure}[ht]
  \centering
  \resizebox{\linewidth}{!}{\input{../figures/single-thread-read-write.tikz}}
  \caption{A single thread reads from and writes to a variable.}
\label{figure:single}
\end{figure}

In Figure~\ref{figure:multiple}, on the other hand, two threads concurrently read and write on $x$. Because the operations of the threads are interleaved, a read operation on $x$ may not yield a value that matches the value of that thread's most recent write. The value may even be inconsistent between consecutive reads. If a single thread incorrectly assumed that it had exclusive control of $x$, this behavior could lead to inconsistent or incorrect results. Similar errors that arise due to an unintentional dependency on execution order are commonly referred to as race conditions.

\begin{figure}[ht]
  \centering
  \resizebox{\linewidth}{!}{\input{../figures/multiple-threads-read-write.tikz}}
  \caption{Two threads concurrently read from and write to a variable.}
\label{figure:multiple}
\end{figure}

Consistency models guarantee the correctness of a program by defining a set of all allowed histories, or histories that lead to a correct result~\cite{Ziv2015}. If an execution of a program follows an allowed history, the execution is consistent. If not, the execution is inconsistent. If every possible execution of the program follows an allowed history, the program conforms to the consistency model.

@@

The examples above assume that all operations complete instantaneously. In real systems, though, operations take time. Even writing to a location in cache memory, an operation measured in nanoseconds, is not truly instantaneous. That an operation may be completed at a time significantly later than the time at which it was invoked introduces uncertainty into a history of operations---the sequence of execution may be influenced by the time it takes for messages to travel.

\begin{figure}[ht]
  \centering
  \resizebox{0.63\linewidth}{!}{\input{../figures/noninstantaneity.tikz}}
  \caption{Two threads concurrently (and noninstantaneously) write to and read from a variable.}
\label{figure:time}
\end{figure}

Consider a third example (Figure~\ref{figure:time}) in which two threads interact with a variable $x$ noninstantanously. Invocation is denoted by the leftmost point of an event and completion by the rightmost point. The dotted line indicates the instant at which the operation takes effect. The lower thread invokes a read while the value of $x$ is set to $1$. The upper thread invokes and completes a write between the time that the read is invoked and the value is actually read. The read operation then returns $2$ even though $x$ was equal to $1$ at the time it was invoked.

While travel time may introduce ambiguity, there are still some restrictions on possible sequences of events. Specifically, an operation cannot take effect before its invocation, nor can it take effect after its completion. Therefore, assuming that all threads share some single global state and that operations on that state take place atomically, each operation will appear to take effect atomically at some point after it is invoked and before it is completed.

\begin{figure}[ht]
  \centering
  \resizebox{0.63\linewidth}{!}{\input{../figures/linearizability.tikz}}
  \caption{A single thread performs an atomic write operation on a variable.}
\label{figure:linearizability}
\end{figure}

This consistency model is referred to as linearizability (also referred to as atomicity or indivisibility), and it guarantees that the completion of an operation on a single object will appear to the rest of the system to be instantaneous~\cite{Herlihy1990}. In other words, even though linearizable operations are executed concurrently and take time, they appear to happen in a simple linear order. Figure~\ref{figure:linearizability} demonstrates an atomic write on a variable $x$: Although the operation takes time, the entire operation (denoted by the shaded area) can be collapsed into one apparently instantaneous event.

A useful consequence of linearizability is that the results of an operation must be visible as soon as the operation is complete. This can prevent issues such as stale reads (in which a read operation returns a value different from the most recently written value) and non-monotonic reads (in which a later read operation returns an older value than an earlier read operation). Linearizability is also a composable guarantee~\cite{Herlihy1990}---an operation made up of smaller linearizable operations is itself linearizable.

@@

\begin{figure}[ht]
  \centering
  \resizebox{0.63\linewidth}{!}{\input{../figures/serializability.tikz}}
  \caption{Two threads perform a set of serializable read and write operations on a variable.}
\label{figure:serializability}
\end{figure}

@@

Serializability alone is a fairly weak consistency model. Because it does not place bounds on time or order, events may happen out of sequence, and, unlike linearizability, serializability is not a composable guarantee. Serializability is useful, though, as a guarantee of isolation: While a serializable set of operations is being executed, it appears to be the only set of operations being executed. For histories to be serializable, then, they must not overlap and cannot interfere with each other.

@@

Atomic variables are used to share independent objects that do not demand coordinated updates but need to be shared by multiple threads~\cite{Swalens2014}. Atomic variables can only be read and mutated by operations that are guaranteed to be linearizable. For example, the atomic operation \code{compare-and-swap} compares the current value of a variable to a given value and only performs a write if the values match~\cite{Swalens2014}. \code{compare-and-swap} ensures that a new value cannot be written based on outdated information. Suppose that a thread $A$ reads a variable $x$ and begins computing a new value. At the same time, another thread $B$ modifies $x$. When $A$ tries to set the value of $x$ via \code{compare-and-swap}, the write will fail. As a result, all writes on $x$ are guaranteed to be linearizable.

@@

Atomic variables often serve as the basis for higher-level concurrency control mechanisms such as semaphores and locks, which in turn support the implementation of concurrent data structures. A counting semaphore is an atomic variable that serves as a record of how many units of a shared resource are available. By taking advantage of atomic operations, the semaphore can be safely updated and will reliably determine whether a shared resource can be used. A single thread can then test if a resource is available and acquire it before proceeding, thereby preventing race conditions~\cite{Swalens2014}. A lock (also referred to as a mutex) guarantees mutual exclusion, which requires that two threads cannot operate on a shared object at the same time. Locks are commonly implemented by a binary semaphore (that is, a semaphore that simply indicates whether a single resource is available).

@@

However, the requirement that transactions must be able to be aborted and retried places constraints on the set of allowed operations. Specifically, a transaction cannot execute any operation that cannot be undone or retried, such as writing to disk or performing a network request. Operations that block indefinitely may also cause a transaction to fail continuously. STM can result in a decrease in performance caused by the overhead of maintaining the log and aborting and retrying transactions.

@@

Other concurrently models avoid the use of shared memory by restricting threads to private memory. Threads communicate strictly by message passing, which avoids issues such as race conditions. Sending messages and reacting to received messages are inherently asynchronous processes, so communicating threads have long been a solution to describing concurrent operations. Limiting thread communication to message passing also enforces modularity and limits possible negative side effects. A wide variety of communicating thread models exist and often run event loops or handle communication with external systems~\cite{Swalens2014}.

@@

Other models, most notably the actor model, rely on asynchronous message passing to specific entities (actors) rather than named channels~\cite{Agha1986}. Actors can only communicate via asynchronous message passing. Upon receiving a message, an actor can choose to send messages to other actors, create new actors, and determine the behavior used for the next received message.

@@

In practice, proxies are used to execute long-running operations such as rendering or network requests~\cite{Swalens2014} without blocking other operations. Promise objects, which can be easily passed and composed, generally offer more flexibility than constructs such as asynchronous callbacks.

@@

While concurrency models guarantee their own correctness, issues may arise when different models are used together. For example, an implementation of an STM that guarantees linearizability assumes that all shared resources are managed by the STM~\cite{Shavit1995}. If such a resource is modified outside of the STM, an unexpected interleaving of operations could result.

Swalens et al.~\cite{Swalens2014} attempted to identify issues that arise when different concurrency models are combined, hoping that future work could identify common ``building blocks'' that could be used to compose a variety of concurrency models. The study surveyed all of the concurrency models available in the Clojure programming language and used them in tandem with one another to see what types of correctness issues were encountered.

@@

Liveness guarantees that ``something good will eventaully happen,'' or that a program will eventually terminate if its input is correct. Taken together, safety and liveness are referred to as total correctness---given a correct input, a program will terminate with the correct output~\cite{Swalens2014}. Deadlocks and livelocks are the primary obstacle to liveness and can arise when models are combined.

@@

To discover the types of conflicts that might arise when combining different models, the study examined all pairwise combinations of Clojure's models. Using every model within every other model uncovered the following types of conflicts:

@@

  \item A model might reexecute code that causes the reexecution to continually happen. The authors of the study observed that multiple STM transactions might consistently conflict with another and never succeed.

@@

The study noted that some types of bad interactions were already prevented by Clojure; for instance, sending a message to an agent from within a transaction is delayed until the transaction succeeds. The authors noted that similar safeguards could prevent other negative interactions. However, preventing every type of negative interaction between models might undermine the purpose of the models to begin with; channels, as mentioned previously, block by design and are inherently incompatible with other models. Therefore, further research would likely seek to decompose concurrency models into common elements and provide a method to compose those elements safely and efficiently.

@@

Marr and D'Hondt~\cite{Marr2012} surveyed a variety of concurrent and parallel programming concepts such as immutability, critical sections, ownership, and common concurrency models to identify a unifying concurrency abstraction that could be implemented in a high-level language virtual machine.\footnote{A 2009 survey by Marr, Haupt, and D'Hondt~\cite{Marr2009} found that virtual machines at the time did not provide intrinsic support for more than one specific model.} A unifying abstraction would enable each concurrency model in a high-level language to be represented in machine code without sacrificing performance or semantics. For instance, a concurrency model in Clojure could be accurately represented by an underlying abstraction in the Java Virtual Machine, and code written in the Java language could interact with that model correctly. To identify the concepts that would need to be supported at the virtual machine level, Marr and D'Hondt selected four questions:

@@

  \item Is the concept already supported by a common virtual machine like the Java Virtual Machine (JVM) or the Common Language Runtime (CLR)?
\end{enumerate}

@@

\begin{itemize}
  \item Managed mutation: A concurrency model may impose rules on how some state can be modified; therefore, mutation must be able to be flexibly managed.

  \item Managed execution: Similarly, invocation of operations on objects is usually restricted and must be able to be flexibly managed.

  \item Ownership: An owning entity regulates mutation and execution, so ownership should be supported in a way that allows for adaptable mutation and execution rules.

@@

  \item Enforceability: All defined restrictions should be enforceable on different concurrency models. Additionally, the semantics of the language in which a model is defined must be enforced regardless of where the model is used.
\end{itemize}

\begin{figure}[ht]
  \centering
  \resizebox{0.63\linewidth}{!}{\input{../figures/obmop.tikz}}
  \caption{Marr and D'Hondt's object-based meta-object protocol~\cite{Marr2012}.}
\label{figure:obmop}
\end{figure}

Given those requirements, Marr and D'Hondt defined an ownership-based meta-object protocol (MOP) that could describe a low-level abstraction, illustrated in Figure~\ref{figure:obmop}. The owner of an object, referred to as the domain, manages operations on all of the objects that it owns (namely, reading and writing of object fields and invocation of methods on objects). This satisfies the ownership requirement. Read and write operations are directed to the \code{readField} and \code{writeField} operations on the domain, satisfying the managed mutation requirement. Similarly, all method invocations are directed to the \code{exec}, satisfying the managed execution requirement. Threads are executed in a domain and specify whether execution occurs at the language level (with restricted reflection) or the meta level (with unrestricted reflection), satisfying the leveled reflection requirement. Globally shared resources are handled by the domain (specifically, the \code{readGlobal} and \code{writeGlobal} methods) if they might break semantics. Additionally, the \code{primitive*} operations allow callers to override the semantics of virtual machine primitives, completing the enforcability requirement.

>>>>

Given that an application is likely to make use of more than one concurrency model, programmers would prefer that different types of models could safely interact. However, different models do not necessarily work well together, nor are they designed to. Recent work has attempted to identify common ``building blocks'' that could be used to compose a variety of models, possibly eliminating subtle problems when different models interact and allowing models to be represented at lower levels without resorting to rough approximations~\cite{Marr2012, Swalens2014, Ziv2015}.

@@

Consistency models, restrictions on possible execution histories, can guarantee that a program will produce a correct result~\cite{Ziv2015}. If an execution of a program follows an allowed history, that execution is said to be consistent; if not, it is said to be inconsistent. If every possible execution of the program is guaranteed to follow a history allowed by a consistency model, that program conforms to that consistency model~\cite{Kingsbury2014}.

\begin{figure}[ht]
  \centering
  \resizebox{0.95\linewidth}{!}{\input{../figures/single-thread-read-write.tikz}}
  \caption{A single thread reads from and writes to a variable.}
\label{figure:single}
\end{figure}

Figure~\ref{figure:single} illustrates a simple program in which a single thread $A$ reads ($=$) and writes ($\leftarrow$) values on a variable $x$, denoted by a segmented bar containing the value of $x$ over time. The program satisfies an intuitive model of how variables should behave---each time $x$ is read, the value returned is equal to the value most recently written and the consistency model holds. If, however, a read on $x$ failed to return the most recently written value, the consistency model would be violated.

\begin{figure}[ht]
  \centering
  \resizebox{0.85\linewidth}{!}{\input{../figures/multiple-threads-read-write.tikz}}
  \caption{Two threads concurrently read from and write to a variable.}
\label{figure:multiple}
\end{figure}

In Figure~\ref{figure:multiple}, two threads $A$ and $B$ concurrently read and write on $x$. Because the operations of the threads are interleaved, a read operation on $x$ may not yield a value that matches the value of that thread's most recent write. The value may even be inconsistent between consecutive reads. If a single thread assumed that it had exclusive control of $x$, incorrectly applying the single variable consistency model, this behavior would appear to be inconsistent. Similar errors that arise due to an unintentional dependency on execution order are commonly referred to as race conditions.

@@

The examples above assume that all operations complete instantaneously. In real systems, though, operations take time. Even writing to a location in cache memory, an operation measured in nanoseconds, is not truly instantaneous. The fact that operations are completed at some time after they are invoked introduces uncertainty into a history of operations---the sequence of execution may be influenced by the time it takes for messages to travel~\cite{Kingsbury2014}.

\begin{figure}[ht]
  \centering
  \resizebox{0.8\linewidth}{!}{\input{../figures/noninstantaneity.tikz}}
  \caption{Two threads concurrently (and noninstantaneously) write to and read from a variable.}
\label{figure:time}
\end{figure}

Consider a third example (Figure~\ref{figure:time}) in which two threads $A$ and $B$ interact with a variable $x$ noninstantanously. Invocation is denoted by the leftmost point of an event and completion by the rightmost point. The dotted line indicates the instant at which the operation takes effect. The lower thread invokes a read while the value of $x$ is set to $2$. The upper thread invokes and completes a write between the time that the read is invoked and the value is actually read. The read operation then returns $3$ even though $x$ was equal to $2$ at the time it was invoked.

While travel time may introduce ambiguity, there are still some restrictions on possible sequences of events. Specifically, an operation cannot take effect before its invocation, nor can it take effect after its completion; rather, it will appear to take effect atomically at some point after it is invoked and before it is completed.

This consistency model is referred to as linearizability (also referred to as atomicity or indivisibility), and it guarantees that the completion of a single operation on a single object will appear to the rest of the system to be instantaneous~\cite{Herlihy1990}. In other words, even though linearizable operations are executed concurrently and take time, they appear to happen in a simple linear order.

\begin{figure}[ht]
  \centering
  \resizebox{0.76\linewidth}{!}{\input{../figures/linearizability.tikz}}
  \caption{A single thread performs an atomic write operation on a variable.}
\label{figure:linearizability}
\end{figure}

Figure~\ref{figure:linearizability} demonstrates an atomic write on a variable $x$: Although the operation takes time, the entire operation (denoted by the shaded area) can be collapsed into one apparently instantaneous event. A useful consequence of linearizability is that the results of an operation must be visible as soon as the operation is complete. This can prevent issues such as stale reads (in which a read operation returns a value different from the most recently written value) and non-monotonic reads (in which a later read operation returns an older value than an earlier read operation). Linearizability is also a composable guarantee~\cite{Herlihy1990}---an operation made up of smaller linearizable operations is itself linearizable.

@@

\begin{figure}[ht]
  \centering
  \resizebox{0.76\linewidth}{!}{\input{../figures/serializability.tikz}}
  \caption{Two threads perform a set of serializable read and write operations on a variable.}
\label{figure:serializability}
\end{figure}

@@

Serializability alone is a fairly weak consistency model because it does not place any restrictions on time or order~\cite{Kingsbury2014}. Events may happen out of sequence, and, unlike linearizability, serializability is not a composable guarantee. Serializability is useful, though, as a guarantee of isolation: While a serializable set of operations is being executed, it appears to be the only set of operations being executed. For histories to be serializable, then, they must not overlap and cannot interfere with each other.

@@

Atomic variables are used to share independent objects that do not demand coordinated updates but need to be shared by multiple threads~\cite{Swalens2014}. Atomic variables can only be read and mutated by operations that are guaranteed to be linearizable. For example, the atomic operation \code{compare-and-swap} compares the current value of a variable to a given value and only performs a write if the values are equal~\cite{Swalens2014}. \code{compare-and-swap} ensures that a new value cannot be written based on outdated information: Suppose that a thread $A$ reads a variable $x$ and begins computing a new value. At the same time, another thread $B$ modifies $x$. When $A$ tries to set the value of $x$ via \code{compare-and-swap}, the write will fail. As a result, all writes on $x$ are guaranteed to be linearizable.

@@

Atomic variables often serve as the basis for higher-level concurrency control mechanisms such as semaphores and locks, which in turn support the implementation of concurrent data structures. A counting semaphore is an atomic variable that serves as a record of how many units of a shared resource, such as a connection or a variable, are available. By taking advantage of atomic operations, the semaphore can be safely updated and will reliably determine whether a shared resource can be used. A single thread can then test if a resource is available and acquire it before proceeding, thereby preventing race conditions~\cite{Swalens2014}. A lock, also referred to as a mutex, guarantees mutual exclusion, which requires that two threads cannot operate on a shared object at the same time. Locks are commonly implemented by a binary semaphore (that is, a semaphore that simply indicates whether a single resource is available).

@@

However, the requirement that transactions must be able to be aborted and retried places constraints on the set of allowed operations. Specifically, a transaction cannot execute any operation that cannot be undone, such as writing to disk or performing a network request. Operations that block indefinitely, like waiting to acquire a lock, may also cause a transaction to fail continuously. STM can also result in a decrease in performance caused by the overhead of maintaining the log and aborting and retrying transactions.

@@

Other concurrently models avoid the use of shared memory by restricting threads to private memory. Threads communicate strictly by message passing, which avoids issues such as race conditions. Sending messages and reacting to received messages are inherently asynchronous processes, so communicating threads have long been a solution to describing concurrent operations. Limiting thread communication to message passing also enforces modularity and limits possible negative side effects. A wide variety of communicating thread models exist and are often used to implement event loops or handle communication with external systems~\cite{Swalens2014}.

@@

Other models, most notably the actor model, rely on asynchronous message passing to specific entities (actors) rather than named channels~\cite{Agha1986}. Actors can only communicate by asynchronous message passing. Upon receiving a message, an actor can choose to perform an operation on a resource that it controls, send messages to other actors, create new actors, determine the behavior used for the next received message, or ignore the message entirely.

@@

In practice, proxies are used to execute long-running operations such as rendering or network requests~\cite{Swalens2014} without blocking other operations. Promise objects, which can be easily passed and composed, offer more flexibility than constructs such as asynchronous callbacks.

@@

Even if concurrency models guarantee their own correctness, issues may arise when different models are used together. For example, an implementation of an STM that guarantees linearizability assumes that all shared resources are managed by the STM~\cite{Shavit1995}. If such a resource is modified outside of the STM, an unexpected interleaving of operations could result.

Swalens et al.~\cite{Swalens2014} attempted to identify issues that arise when different concurrency models are combined, hoping that future work could identify common ``building blocks'' that could be used to compose a variety of concurrency models. The study surveyed all of the concurrency models available in the Clojure programming language and used them in within one another to see what types of correctness issues were encountered.

@@

Liveness guarantees that ``something good will eventaully happen,'' or that a program will eventually terminate if its input is correct. Taken together, safety and liveness are referred to as total correctness---given a correct input, a program will terminate with the correct output~\cite{Swalens2014}. Deadlocks (in which execution is blocked) and livelocks (in which execution continues indefinitely but never makes progress) are the primary obstacle to liveness and can arise when models are combined.

@@

To discover the types of conflicts that might arise when combining different models, the study examined all pairwise combinations of Clojure's models. Using every model within every other model (for example, manipulating an atomic variable within an STM transaction) uncovered the following types of conflicts:

@@

  \item A model might reexecute code that causes the reexecution to continually happen. The authors of the study observed that, especially when many transactions are being executed at once, a large STM transaction might consistently conflict with another and never succeed.

@@

The study noted that some types of bad interactions are prevented by Clojure; for instance, sending a message to an agent from within a transaction is delayed until the transaction succeeds. The study also noted that similar safeguards could prevent other negative interactions. However, preventing every type of negative interaction between models might undermine the purpose of the models to begin with; channels, as mentioned previously, block by design and are inherently incompatible with other models. With that in mind, further research would likely seek to decompose concurrency models into common elements and provide a method to compose those elements safely and efficiently.

@@

Marr and D'Hondt~\cite{Marr2012} surveyed a variety of concurrent and parallel programming concepts such as immutability, critical sections, ownership, and common concurrency models to identify a unifying concurrency abstraction that could be implemented in a high-level language virtual machine. A unifying abstraction would enable each concurrency model in a high-level language to be represented in machine code without sacrificing performance or semantics. For instance, a concurrency model in Clojure could be accurately represented by an underlying abstraction in the Java Virtual Machine, and code written in the Java language could interact with that model correctly. To identify the concepts that would need to be supported at the virtual machine level, Marr and D'Hondt selected four questions:

@@

  \item Is the concept already supported by a common virtual machine like the Java Virtual Machine or the Common Language Runtime?
\end{enumerate}

@@

\begin{itemize}
  \item Managed mutation and execution: A model may impose rules on how objects can be modified; therefore, mutation must be handled in a way that allows those rules to be enforced. Similarly, a model might restrict the invocation of operations on objects, and those rules must be enforced. For example, a thread should not be permitted to mutate a variable that a model declares immutable.

  \item Ownership: All mutation an execution on an object is regulated relative to some ``owning'' entity, so ownership should be supported in a way that allows for adaptable mutation and execution rules.

@@

  \item Enforceability: All defined restrictions should be enforceable on different concurrency models. Additionally, the semantics of the language (immutability, for example) in which a model is defined must be enforced regardless of where the model is used.
\end{itemize}

\begin{figure}[ht]
  \centering
  \resizebox{0.55\linewidth}{!}{\input{../figures/obmop.tikz}}
  \caption{Marr and D'Hondt's object-based meta-object protocol~\cite{Marr2012}.}
\label{figure:obmop}
\end{figure}

Given those requirements, Marr and D'Hondt defined an ownership-based meta-object protocol (MOP) that could describe a low-level abstraction, illustrated in Figure~\ref{figure:obmop}. The owner of an object, referred to as the domain, manages operations on all of the objects that it owns. Specifically, the domain handles all reading and writing of object fields and invocation of methods on objects. This satisfies the ownership requirement. Read and write operations are directed to the \code{read\_field} and \code{write\_field} operations on the domain, satisfying the managed mutation requirement. Similarly, all method invocations are directed to the \code{exec}, satisfying the managed execution requirement. Threads are executed in a domain and specify whether execution occurs at the language level with restricted reflection or the meta level with unrestricted reflection, satisfying the leveled reflection requirement. Globally shared resources are handled by the domain (through the \code{read\_global} and \code{write\_global} methods) if they might break semantics. Additionally, the \code{primitive*} operations allow callers to override the semantics of virtual machine primitives, completing the enforcability requirement.
